---
title: "Activity monitoring"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data processing and setup

First of all, we will read the data into R which was provided in the forked repository [retrieved 2020 07 12].
Data can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip.
```{r reading, cache=TRUE}
     data<-read.csv(unz("activity.zip", "activity.csv"), na.strings = "NA")       
```

Next, we will load the packages that will be used for analysis:

```{r package, message=FALSE}
library(dplyr)
library(ggplot2)
```


## Data Analysis & Results
  
### Steps taken per day

First of all, we will create a data frame, that finds the sum of steps taken each day

```{r stepday, message=FALSE}
stepday <- data %>% group_by(date = as.factor(date)) %>% 
        summarize(stepsum = sum(steps, na.rm=T))

head(stepday)
```

Histogram of the total number of steps taken each day:

```{r histogram}
hist(stepday$stepsum, breaks=8, xlab="Steps", 
     main="Histogram of Steps Per Day", col="azure")
```

Here is the mean and median of steps taken per day

```{r meanmedian}
data.frame(Mean = mean(stepday$stepsum), Median = median(stepday$stepsum))
```

### Average dayly activity pattern

First, we will group the data by intervals and take the average of each interval:

```{r avgstep, message=FALSE}
avgstep <- data %>% group_by(interval) %>% summarize (average = mean(steps, na.rm=T))

head(avgstep)
```

Next, we will plot the this average over each interval:

```{r avgstepplot}
ggplot(data = avgstep, aes(interval, average)) + 
        geom_line(colour="blue") + 
        labs(title="Average Steps", subtitle ="In Interval Time")
```

The interval that has the highest step average is `r avgstep[which(avgstep$average == max(avgstep$average)), "interval"]`

### Imputing missing values

Total number of rows with NA's is `r length(!complete.cases(data))`

For each missing case, we used the mean for that 5-minute interval to fill in the missing data:

```{r filldata}
fill<-data
for (i in seq_along(fill$steps))
{
      if(is.na(fill$steps[i]))
        {
                fill$steps[i]<-as.numeric(avgstep[which(fill$interval[i] ==
                avgstep$interval),"average"])
        }
}
```

A new data frame was created to group the filled data by dates and calculate the sum over each day:

```{r stepday2, message=FALSE}
stepday2 <- fill %>% group_by(date = as.factor(date)) %>% 
        summarize(stepsum = sum(as.numeric(steps)))
```

For a better comparison, two graphs were drawn (the first one - with NA, the second - without)

```{r histogram2}
par(mfrow=c(1,2))

hist(stepday$stepsum, breaks=8, xlab="Steps", 
     main="Histogram of Steps Per Day", col="azure")

hist(stepday2$stepsum, breaks = 8, xlab="Steps", 
     main="Histogram of Steps Per Day (No NA)", col="azure")

par(mfrow=c(1,1))
```

The mean and median with the missing values filled in:

```{r meanmed2}
data.frame(Mean = c(mean(stepday$stepsum), mean(stepday2$stepsum)), 
           Median = c(median(stepday$stepsum),median(stepday2$stepsum)),
           row.names=c("with NA", "no NA"))
```

### Weekdays and Weekends

#### Dividing days into weekdays and weekends

Firstly, the character variable "date" of the dataset was converted into the R date variable.

```{r replace1}
fill$date <- as.Date(fill$date, "%Y-%m-%d")
```

Secondly, a new variable was introduced into the dataset, giving all the row values the name "Weekday":

```{r replace2}
fill$Day <- "Weekday"
```

Lastly, using the weekdays() function, the weekend values were replaced to have the name "Weekend"

```{r replace3}
fill$Day <- as.factor(replace(fill$Day, which(weekdays(fill$date) == 
            "Saturday" | weekdays(fill$date) == "Sunday"), "Weekend"))
```

#### Differences

The new data vas summarized by taking the interval means.
```{r avgstepw, message=FALSE}
avgstepw <- fill %>% group_by(interval, Day) %>% 
        summarize (average = mean(as.numeric(steps)))
```

A graph was plotted, subseting the data over Weekdays and Weekends
```{r avgstepwplot}
ggplot(data = avgstepw, aes(interval, average,colour = Day)) + 
        geom_line(show.legend=F) +
        facet_grid(Day~.) +
        labs(title="Average Step count for each time interval during weekdays and weekends")
```

As we can see, the activity during the weekdays and weekends seems to be similar. The movement during the weekends is more consistent, whereas during the weekdays there is a spike of movement in the ~850th interval.
